Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-22 paper.qmd
Update paper.qmd

However, [renv]{.pkg} has important limitations. It does not restore the version
of [R]{.proglang} itself—installing the correct R version must be done
separately using tools like [rig]{.pkg} [@rlib2023]. More critically,
[renv]{.pkg} does not handle system-level dependencies. If [sf]{.pkg} requires
[GDAL]{.pkg} version 3.0 but the system has version 2.4 installed, [renv]{.pkg} cannot
resolve this conflict. Users must manually install system libraries, and the
process differs across operating systems. Further details are discussed in the
[renv]{.pkg} documentation:
\url{https://rstudio.github.io/renv/articles/renv.html#caveats}.

@@

```{=tex}
\begin{CodeInput}
$> nix-shell -I \
  nixpkgs=https://github.com/rstats-on-nix/nixpkgs/archive/refs/heads/2025-10-17.tar.gz -p \
  R rPackages.rix
\end{CodeInput}
```

@@

[rix]{.pkg} can also handle more complex setups:

```{=tex}
\begin{CodeInput}
R> rix(r_ver = "4.3.1",
+    r_pkgs = c("dplyr", "chronicler", "AER@1.2-8"),
+    system_pkgs = c("quarto", "git"),
+    tex_pkgs = c(
+          "amsmath",
+          "framed",
+          "fvextra",
+          "environ",
+          "fontawesome5",
+          "orcidlink",
+          "pdfcol",
+          "tcolorbox",
+          "tikzfill"
+    ),
+    git_pkgs = list(
+      list(
+        package_name = "rix",
+        repo_url = "https://github.com/b-rodrigues/rix/",
+        branch_name = "master",
+        commit = "ea92a88ecdfc2d74bdf1dde3e441d008521b1756"),
+      list(
+        package_name = "fusen",
+        repo_url = "https://github.com/ThinkR-open/fusen",
+        branch_name = "main",
+        commit = "d617172447d2947efb20ad6a4463742b8a5d79dc")
+    ),
+    ide = "positron",
+    project_path = ".",
+    overwrite = TRUE)
\end{CodeInput}
```

This call to `rix()` generates a `default.nix` that builds a development shell
which ships [R]{.proglang} and [R]{.proglang} packages (including [AER]{.pkg} version 1.2-8), several TeXLive
packages for \LaTeX document authoring, development versions of [rix]{.pkg} and
[fusen]{.pkg} from GitHub, and the Positron editor.

@@

To illustrate these capabilities with a realistic research task, we present a
polyglot pipeline that simulates a Real Business Cycle (RBC) model in
[Julia]{.proglang} [@bezanson2017julia], uses the resulting data to train an
[XGBoost]{.pkg} forecasting model in [Python]{.proglang}, visualizes the results
in [R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a [Quarto]{.pkg}
report.

@@

For Python users, a port called [ryxpress]{.pkg} allows building the same
pipelines and inspecting outputs from Python sessions. [rixpress]{.pkg} also
includes several additional features not covered here for brevity. It is also
possible to configure popular IDEs to work interactively and seamlessly with
both [rix]{.pkg} and [rixpress]{.pkg}, enabling a smooth, reproducible workflow
from within the development environment. Detailed setup instructions are
provided in the vignettes of both packages.

@@

We thank the rOpenSci reviewers and contributors who provided valuable feedback
on the development of [rix]{.pkg} and [rixpress]{.pkg}. In particular, we are
grateful to David Watkins and Jacob Wujiciak-Jens for their reviews of `{rix}`,
and to William Landau and Anthony Martinez for their reviews of `{rixpress}`. We
also acknowledge the contributions of Richard J. Acton, Jordi Rosell, Elio
Campitelli, László Kupcsik, and Michael Heming for `{rix}`. Their expertise and
feedback greatly improved the quality and usability of these packages.

@@

## Appendix: A Complete Polyglot Example with rixpress {#sec-appendix}

@@

### The Environment Definition

@@

### The RBC Model Simulation

@@

### The XGBoost Forecasting Model

>>>>

However, [renv]{.pkg} has important limitations. It does not restore the version
of [R]{.proglang} itself—installing the correct R version must be done
separately using tools like [rig]{.pkg} [@rlib2023]. More critically,
[renv]{.pkg} does not handle system-level dependencies. If [sf]{.pkg} requires
[GDAL]{.pkg} version 3.0 but the system has version 2.4 installed, [renv]{.pkg} cannot
resolve this conflict. Users must manually install system libraries, and the
process differs across operating systems. Further details are discussed in the
[renv]{.pkg} 
documentation^[\url{https://rstudio.github.io/renv/articles/renv.html#caveats}].

@@

```{=tex}
\begin{CodeInput}
$> nix-shell -I \
+  nixpkgs=https://github.com/rstats-on-nix/nixpkgs/archive/2025-10-20.tar.gz -p \
+  R rPackages.rix
\end{CodeInput}
```

@@

[rix]{.pkg} can also handle more complex setups, and users can provide a date
instead of a specific [R]{.proglang} version:

```{=tex}
\begin{CodeInput}
R> rix(date = "2025-10-20",
+    r_pkgs = c("rix", "dplyr", "chronicler", "AER@1.2-8"),
+    system_pkgs = c("quarto", "git"),
+    tex_pkgs = c(
+          "amsmath",
+          "framed",
+          "fvextra",
+          "environ",
+          "fontawesome5",
+          "orcidlink",
+          "pdfcol",
+          "tcolorbox",
+          "tikzfill"
+    ),
+    git_pkgs = list(
+        package_name = "fusen",
+        repo_url = "https://github.com/ThinkR-open/fusen",
+        commit = "60346860111be79fc2beb33c53e195f97504a667"
+    ),
+    ide = "positron",
+    project_path = ".",
+    overwrite = TRUE)
\end{CodeInput}
```

This call to rix() generates a `default.nix` file for a development shell that encapsulates 
a complete and reproducible research environment. It provides [R]{.proglang} and its packages 
(including [AER]{.pkg} at version 1.2-8), several TeXLive packages for \LaTeX document authoring, 
development versions of [rix]{.pkg} and [fusen]{.pkg} pulled directly from GitHub at a specific commit,
and the Positron editor. The reproducibility of this environment is guaranteed by pinning all 
components to a single point in time: the [R]{.proglang} packages are resolved from the CRAN 
snapshot of October 20, 2025, while all other system tools are fixed to the version of 
`nixpkgs` from that same date. Typing `nix-shell` in a terminal in the folder that containts
this `default.nix`, will drop the user into the shell. It should be noted that it is 
also possible to configure IDEs to dynamically load this shell to provide the right
tooling for developing.

@@

To illustrate these capabilities with a realistic research task, we present a
polyglot pipeline that simulates a Real Business Cycle (RBC) model in
[Julia]{.proglang} [@bezanson2017julia], uses the resulting data to train an
[XGBoost]{.pkg} forecasting model in [Python]{.proglang}, visualizes the results
in [R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a [Quarto]{.pkg}
report. The code to this example can be found on 
[following repository](https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc)^[https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc],
with additional details provided in the Appendix of this manuscript.

@@

For Python users, a port called [ryxpress]{.pkg} allows building the same
pipelines and inspecting outputs from Python sessions. [rixpress]{.pkg} also
includes several additional features not covered here for brevity. As mentioned
previously, it is also possible to configure popular IDEs to work interactively and seamlessly with
both [rix]{.pkg} and [rixpress]{.pkg}, enabling a smooth, reproducible workflow
from within the development environment. Detailed setup instructions are
provided in the vignettes of both packages.

@@

I thank the rOpenSci reviewers and contributors who provided valuable feedback
on the development of [rix]{.pkg} and [rixpress]{.pkg}. In particular, I am
grateful to David Watkins and Jacob Wujiciak-Jens for their reviews of `{rix}`,
and to William Landau and Anthony Martinez for their reviews of `{rixpress}`. I
also acknowledge the contributions of Philipp Baumann, [rix]{.pkg}’s co-author,
Richard J. Acton, Jordi Rosell, Elio Campitelli, László Kupcsik, and 
Michael Heming for `{rix}`. Their expertise and feedback greatly 
improved the quality and usability of these packages.

@@

## Appendix {#sec-appendix}

### Reproducing this paper

The source code of this paper is hosted on GitHub and can be found
on following 
[link](https://github.com/b-rodrigues/rix_paper)^[https://github.com/b-rodrigues/rix_paper].
The paper can be easily compiled by running the following command:

```{=tex}
\begin{CodeInput}
$> nix-shell --run "quarto render paper.qmd --to jss-pdf"
\end{CodeInput}
```

This command utilizes the `default.nix` file to build the exact computational 
environment required to compile the manuscript. To ensure continuous reproducibility,
the paper is also automatically re-compiled via GitHub Actions on every push to the 
repository, using the same Nix environment.

### A Complete Polyglot Example with rixpress

@@

It is possible to run this complete example by executing the shell
script called `run_polyglot_example.sh` found in the repository of
this 
paper^[\url{https://github.com/b-rodrigues/rix_paper/blob/master/run_polyglot_example.sh}].

#### The Environment Definition

@@

#### The RBC Model Simulation

@@

#### The XGBoost Forecasting Model
